---
title: "Week 11 Lab - Explore Categoical Variable Associations"
subtitle: "*PSY 150* - The Chi-Square Test of Indepenedence"
author: "KEY"
date: "`r Sys.Date()`"
output: 
    html_document
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
options(scipen = 999)
```

------------------------------------------------------------------------

### Outline 

- Example demo (follow-along): Investigate associations between categorical variables
- Lab Activity: Complete coding activity & answer questions exploring categorical variable associations

------------------------------------------------------------------------

Load packages 
```{r}

library(tidyverse)
library(here)
library(janitor)
library(gt)
library(gtsummary)
library(gtExtras)
library(praise)
library(beepr)

```

------------------------------------------------------------------------

### Read in the Simulated data

------------------------------------------------------------------------

```{r}
set.seed(2025)

psy150_data <- read_csv(here("data", "psy150_survey_data.csv")) %>% 
    mutate(across(c(fav_season:birth_month, risk_pref, study_sound), as.factor)) %>% 
    mutate(wake_up_time = factor(wake_up_time, 
      levels = c("6 am", "7 am", "8 am", "9 am", "10 am", "11 am", "12 pm", "1 pm (or later)"))) %>%
   mutate(wake_up_time = as.numeric(wake_up_time)+5) %>% 
   mutate(risk_pref = factor(risk_pref, 
      levels = c("A sure $10", "A 50% chance to win $24"),
      labels = c("Low Risk", "High Risk"))) %>% 
   mutate(random_groups = sample(rep(c("control", "treatment"), each = n()/2)))

```

------------------------------------------------------------------------

### Are the categorical variables `risk_pref` & `random_groups` related (or independent)?

------------------------------------------------------------------------

### Make a Contingency Table using `tbl_summary()`


```{r}
chi_table <- psy150_data %>% 
    select(random_groups, risk_pref) %>% 
    tbl_summary(by = random_groups)

chi_table
```

------------------------------------------------------------------------

### Add Chi-Square Test Result to Table

Include Chi-Square test result (*p-value*) indicating `independence` or `non-independence`
```{r}

chi_table %>% add_p()

```

------------------------------------------------------------------------

### Contingency Table with Marginal Totals

Create contingency table using `tabyl()` & add marginal totals with `adorn_totals()`
```{r}

psy150_data %>% 
tabyl(risk_pref, random_groups) %>% 
  adorn_totals(where = c("row", "col")) %>% 
  gt()
 
```

------------------------------------------------------------------------

### Compare *expected frequencies* (assuming independence) with the *observed frequencies* 

- Null Hypothesis (H0; independence): Variables are unrelated
- Alternative Hypothesis (H1: non-independence): Variables are related
- Expected Frequencies: What we would expect if the 2 variables are independent 
- Chi-Square Test (purpose): To determine whether the *observed freq.* provide evidence of an association (or not)

```{r}

chi_results <- psy150_data %>% 
  tabyl(risk_pref, random_groups) %>% 
  chisq.test(correct = FALSE)
  
tbl_obs <- chi_results$observed %>% 
  gt() %>%
  gt::tab_header(title = "observed")

tbl_exp <- chi_results$expected %>% 
  gt() %>% 
  gt::tab_header(title = "expected")

gt_two_column_layout(list(tbl_obs, tbl_exp))

```

------------------------------------------------------------------------

### Estimate the Chi-Square Test Statisitic using `chisq.test()`

`chisq.test()` returns 3 important parts of the Chi-Square Test:

1. Chi-Square Statistic Estimate (`X-squared`)
2. Degrees of Freedom (`df`): $df = (RowNumber -1)*(ColumnNumber - 1)$
3. P-Value Estimate (Critical value determined by `alpha = .05` and `df = 1`) 

```{r}

psy150_data %>% 
  tabyl(risk_pref, random_groups) %>% 
  chisq.test(correct = FALSE)
  
```

------------------------------------------------------------------------

### Lab Activity - Exploring Categorical Variable Relations

- Step 1: Read in the data needed for each example
- Step 2: Complete the coding task described in the directions 
- Step 3: Read questions after each coding task → Interpret output & respond to questions

------------------------------------------------------------------------

### Example 1: Anxiety During COVID by Occupation

**Study source:** Kush et al. (2022) *Teachers’ Mental Health During the COVID-19 Pandemic*

Read in the simulated teacher mental health data:
```{r}

teachr_data <- read_csv(here("data", "teachers_anx_covid_sim.csv")) %>% 
  mutate(across(c(occupation:occupation2), as.factor)) %>% 
  mutate(anxiety = factor(anxiety, levels = c(0,1), labels = c("Low Anxiety", "High Anxiety"))) %>%
  mutate(occupation2 = factor(occupation2, levels = c("all_others","teachers"),
         labels = c("Other Occupations", "Teachers"))) 

```

------------------------------------------------------------------------

### Are categorical variables `anxiety` & `occupation2` related? 

- `anxiety` (factor levels): `Low Anxiety`, `High Anxiety`
- `occupation2` (factor levels): `Other Occupations`, `Teachers`

------------------------------------------------------------------------

### Create a contingency table with marginal totals

- Use functions: `tabyl()` + `adorn_totals()` + `gt()`
- HINT: Find the corresponding code example above → Update with new data & variables 

```{r}

teachr_data %>% 
  tabyl(anxiety, occupation2) %>% 
  adorn_totals(where = c("row", "col")) %>% 
  gt()

```

**Question:**  What percent of the sample are `Teachers` & what percent of the sample have `Other Occupations`?

**Response:** __________________________________

**Question:**  What percent of `Teachers` reported `Low Anxiety`? [AND] What percent of `Teachers` reported `High Anxiety`?

**Response:** __________________________________

------------------------------------------------------------------------

### Estimate the critical value for this chi-square test (i.e., the significance threshold)

- Choose level of significance (`alpha`) & use function `qchisq()` to determine the `critical value`
- HINT: The `qchisq()` function requires 2 inputs `alpha` and `df`(*degrees of freedom*). 

```{r}
# Choose alpha (level of significance) 
alpha <- .05

qchisq(p = 1-alpha, df = 1)
```

**Question:**  Determine the *degrees of freedom* for a chi-square test using the contingency table above:

**Response:** __________________________________

**Question:** If the estimated `chi-square statistic` is **greater than** the `critical value` what does this tell us about the association between the two categorical variables? [AND]

What does it mean when `chi-square statistic` is **less than** the `critical value`?

**Response:** __________________________________

------------------------------------------------------------------------

### Estimate the Chi-Square test of independence for `anxiety` and `occupation2`

- Use functions: `tabyl()` + `chisq.test()`
- HINT: Find the corresponding code example above → Update with new data & variables 

```{r}

teachr_data %>% 
  tabyl(anxiety, occupation2) %>% 
  chisq.test(correct = FALSE)

```

**Question:**  What does the result of the test tell us about association between the two categorical variables? 

**Response:** __________________________________


------------------------------------------------------------------------

### Example 2: Are the categorical variables `dropped_out` & `gender` related?

**Study source:** Nakajima et al. (2012) *Student Persistence in Community Colleges*

- `dropped_out` (factor levels): `Enrolled`, `Dropped Out`
- `gender` (factor levels): `Female`, `Male`

------------------------------------------------------------------------

Read in the simulated city college persistence data:
```{r}

cc_data <- read_csv(here("data", "city_college_persistence_sim.csv")) %>% 
  mutate(across(c(gender, race, financial_aid), as.factor)) %>%
  mutate(dropped_out = factor(dropped_out, labels = c("Enrolled","Dropped Out")))
  
```

------------------------------------------------------------------------

### Create a contingency table

Use functions: `tabyl()` + `gt()`
```{r}

cc_data %>% 
  tabyl(dropped_out, gender) %>% 
  gt()

```

**Question:**  What percent of female students dropped out? [AND] What percent of male students dropped out?

**Response:** __________________________________

Estimate the Chi-Square Test of Independence 
```{r}

cc_data %>% 
  tabyl(dropped_out, gender) %>% 
  chisq.test(correct = FALSE)

```

**Question:**  What does the result of the test tell us about association between the two categorical variables? 

**Response:** __________________________________

------------------------------------------------------------------------

### Example 3: Are the categorical variables `ethnicity` & `ses_low` related?

**Study source:** Kazeem et al. (2020) *Adverse Childhood Experiences, Socio-Economic Status, and Criminal Behaviour: a Cross-Sectional Correctional Survey*

- `ethnicity` (factor levels): `Hausa`, `Igbo`, `Yoruba`, `Other`
- `ses_low` (factor levels): `High (SES)`, `Low (SES)`

------------------------------------------------------------------------

Read in simulated data on Adverse Childhood Experiences (ACEs) and crime 
```{r}
ace_data <- read_csv(here("data", "ACEs_violent_crime_SIM.csv")) |>
  mutate(
   ethnicity = factor(ethnicity,
      levels = c("hausa","igbo","yoruba","other"),
      labels = c("Hausa","Igbo","Yoruba","Other"))) %>%
  mutate(ses_low = factor(ses_low, labels = c("High (SES)", "Low (SES)"))) 
```

### Create a contingency table

Use functions: `tabyl()` + `gt()`
```{r}

ace_data %>% 
  tabyl(ethnicity, ses_low) %>% 
  gt()

```

**Question:**  Calculate the chi-square test degrees of freedom (`df`) for a contingency table of this size: 

**Response:** __________________________________

**Question:**  What percent of the sample in the `Hausa` ethnicity group reported `Low SES`? [AND] 

What percent of the sample in the `Other` ethnicity group reported `Low SES`?

**Response:** __________________________________

Estimate the Chi-Square Test of Independence 
```{r}

ace_data %>% 
  tabyl(ethnicity, ses_low) %>% 
  chisq.test(correct = FALSE)

```

**Question:**  What does the result of the test tell us about association between the two categorical variables? 

**Response:** __________________________________


------------------------------------------------------------------------

```{r}

praise("${EXCLAMATION} - You have made it to the end of lab!!! You have ${created} something truely ${adjective}!"); beep(1)

```

------------------------------------------------------------------------
