---
title: "Week 9 - Logistic Regression"
subtitle: "PSY 150"
author: "KEY"
date: "`r Sys.Date()`"
output: 
    html_document
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, eval=TRUE)
```

### Load packages

```{r}

library(tidyverse)
library(here)
library(janitor)
library(report)
library(jtools)
library(sjPlot)

source("load_plot_functions.R")

```

### Read in the Simulated data

```{r}

teachr_data <- read_csv(here("data", "teachers_anx_covid_sim.csv")) %>% 
    mutate(across(c(occupation:occupation2), as.factor)) %>% 
    mutate(occupation = factor(occupation,
           levels = c("teachers","healthcare","office","other")))

```

------------------------------------------------------------------------

Categorical outcome variable (`anxiety`): 

0 = Low Anxiety (`none / some of the time`) 
1 = High Anxiety (`most / all of the time`)

Categorical predictor (`occupation2`; factor with 2 levels):

0 = `all other occupations`
1 = `teachers`

------------------------------------------------------------------------

Check the distribution of outcome variable & predictor variable:
```{r}

tabyl(teachr_data$anxiety)
tabyl(teachr_data$occupation2)

```

Plot the categorical outcome & view response proportions for each category:
```{r}

bar_plot(
  data = teachr_data,
  x_var = anxiety,
  x_label = "Anxiety Group (0 = Low Anxiety , 1 = High Anxiety) ",
  title = "Distribution of Categorical Outcome Variable Anxiety "
)

```

------------------------------------------------------------------------

### Specifying a Logistic Regression model

------------------------------------------------------------------------

```{r}
model0 <- glm(anxiety ~ occupation2, 
         data = teachr_data, family = binomial())

summ(model0,
     exp = TRUE, # Change to `FALSE` for Logit Estimates
     model.fit = FALSE, digits = 3)
```

### Why is the Odds Ratio labeled "exp(Est.)" 

```{r}
# To convert logistic regression coefficients estimated 
# in the Log-Odds (Logits) to Odds Ratio (OR):

# EQUATION: OR = e^(logit_beta)
exp(0.318)

```

**Question:**  Why is the Odds Ratio labeled "exp(Est.)" 

**Response:** __________________________________

------------------------------------------------------------------------

### How do we interpret the intercept?

1. Convert the intercept (odds) to probability:

**OR-to-Probability Equation:**  Probability = OR_intercept / (1 + OR_intercept)

2. The intercept is then interpreted as the probability that *Y = 1* predicted by the model when *all predictors (Xs) are held at zero (or the reference group)*:

```{r}
# The intercept probability = 
0.081/(1+0.081)

```

**Question:** What is the interpretation of intercept for this model:

**Response:** __________________________________

------------------------------------------------------------------------

### How do we interpret the predictor?

1. Convert the predictor (odds ratio) to *percent change*:

**OR-to-Percent Equation:**  `Percent Change = (OR-1)*100`

2. The predictor is then interpreted as the percent change in the odds of *Y = 1* predicted by the model when *the predictor X increases by 1 (or for the non-reference group relative to reference group)*:

```{r}
# Predictor (percent change in the odds) = 
(1.375-1)*100

```

**Question:** What is the interpretation of predictor for this model:

**Response:** __________________________________

------------------------------------------------------------------------

### Forward Stepwise Regression: 

**Compare models after adding additional predictors:**

1. Start with focal predictor only and estimate model (`Step 1 Model`)
2. Estimate a new model(`Step 2 Model`) with the addition of a control variable
3. Compare `Step 1 Model` model to `Step 2 Model`
4. Add an additional control to the `Step 2 Model` and estimate `Step 3 Model`...
5. Repeat (2-3): Add additional control variables to previously fitted model. 
   Test all potentially relevant control variables.

------------------------------------------------------------------------

Is the predictor variable continuous or categorical?
```{r}

tabyl(teachr_data$occupation)

```

Categorical predictor (`occupation`; factor with 4 levels):

0 = `teachers` 
1 = `healthcare`
2 = `office`
3 = `other`

Fit the step-1 model (focal predictor only): 
```{r}

mstep1 <- glm(anxiety ~ occupation,
         data = teachr_data, family = binomial())

summ(mstep1, exp = TRUE, model.fit = FALSE, digits = 3)

```

Fit the step-2 model (add control predictor `gender`): 
```{r}

mstep2 <- glm(anxiety ~ occupation + gender,
         data = teachr_data, family = binomial())

summ(mstep2, exp = TRUE, model.fit = FALSE, digits = 3)

```

Fit the step-3 model (add control predictor `age_band`): 
```{r}

mstep3 <- glm(anxiety ~ occupation + gender + age_band,
         data = teachr_data, family = binomial())

summ(mstep3, exp = TRUE, model.fit = FALSE, digits = 3)

```


------------------------------------------------------------------------

### City College Persistence (Predicting dropout)

> Data was simulated to approximately replicate findings from study:

Nakajima et al. (2012): "Student Persistence in Community Colleges"

------------------------------------------------------------------------

Read in the simulated city college persistence data:
```{r}

cc_data <- read_csv(here("data", "city_college_persistence_sim.csv")) %>% 
  mutate(across(c(gender, race, financial_aid), as.factor)) %>%
  mutate(dropped_out = factor(dropped_out, labels = c("Enrolled","Dropped Out")) )
  
```

Plot the categorical outcome & view response proportions for each category:
```{r}

bar_plot(
  data = cc_data,
  x_var = dropped_out,
  x_label = "Student Drop Out Indicator",
  title = "Distribution of Categorical Outcome Variable Student Drop Out"
)

```

------------------------------------------------------------------------

### Logistic Regression with a Continuous Predictor 

```{r}

mstep0_demo <- glm(dropped_out ~ faculty_interaction,
                 data = cc_data, family = binomial())

summ(mstep0_demo, exp = TRUE, model.fit = FALSE, model.info = FALSE, digits = 3)
```

In logistic regression, the change in odds is `multiplicative`, `not additive`:

Equation (Take the OR estimate to the power of X): `OR_BETA^(X)` 

How much does the model predict the odds of `dropping out` changes when `faculty_interaction` changes from `1 to 3`?

Calculate predictor as percent change when `faculty_interaction = 1`: 

**Response:** _______16.9% lower odds of dropping out ________

Calculate predictor as percent change when `faculty_interaction = 3`: 

**Response:** _________42.6% lower odds of dropping out_____


------------------------------------------------------------------------

### GROUP 1: Estimate the regression of the outcome predicted by focal predictor

**Instructions:**

1. Estimate the following logistic regression model:

*dropped_out = ⍺ + β(offcampus_work_hours)*

2. Convert the intercept from Odds-Ratio (OR) to probability and interpret:

**OR-to-Probability Equation:**  Probability = OR_intercept / (1 + OR_intercept)

3. Interpret the Odds Ratio of the predictor `offcampus_work_hours` as percent change

**OR-to-Percent-Change Equation:** Percent Change = (1 - OR)*100

4. Is the predictor estimate for `offcampus_work_hours` significant? 

------------------------------------------------------------------------


```{r}

mstep1_persist <- glm(dropped_out ~ offcampus_work_hours,
                 data = cc_data, family = binomial())

summ(mstep1_persist, exp = TRUE, model.fit = FALSE, model.info = FALSE, digits = 3)
```

Interpretation of intercept (probability): 

**Response:** _______________________

Interpretation of focal predictor `offcampus_work_hours`: 

**Response:** _______________________

Significance of predictor estimate: 

**Response:** _______________________


------------------------------------------------------------------------

### GROUP 2: Add the control variable `cumulative_gpa` to the model

**Instructions:**

1. Estimate the following logistic regression model:

*dropped_out = ⍺ + β1(offcampus_work_hours) + β2(cumulative_gpa)*

2. Interpret the Odds Ratio of the predictor `cumulative_gpa` as percent change

**OR-to-Percent-Change Equation:** Percent Change = (1 - OR)*100

3. Is the predictor estimate for `cumulative_gpa` significant? 

4. How did the addition of the control variable to the model affect the focal predictor est/sig? 

------------------------------------------------------------------------

```{r}
mstep2_persist <- glm(dropped_out ~ 
                   offcampus_work_hours +
                   cumulative_gpa,
                 data = cc_data, family = binomial())

summ(mstep2_persist, exp = TRUE, model.fit = FALSE, model.info = FALSE, digits = 3)

```

Interpretation of control predictor `cumulative_gpa`: 

**Response:** _______________________

Significance of predictor estimate: 

**Response:** _______________________

How did the addition of the control variable to the model affect the focal predictor est/sig? 

**Response:** _______________________

------------------------------------------------------------------------

### GROUP 3: Add the control variable `enrollment_units` to the model

**Instructions:**

1. Estimate the following logistic regression model:

*dropped_out = ⍺ + β1(offcampus_work_hours) + β2(cumulative_gpa) + β3(enrollment_units)*

2. Interpret the Odds Ratio of the predictor `enrollment_units` as percent change

**OR-to-Percent-Change Equation:** Percent Change = (1 - OR)*100

3. Is the predictor estimate for `enrollment_units` significant? 

4. How did the addition of the control variable to the model affect the focal predictor est/sig? 

------------------------------------------------------------------------

```{r}
mstep3_persist <- glm(dropped_out ~ 
                   offcampus_work_hours +
                   cumulative_gpa +
                   enrollment_units,
                 data = cc_data, family = binomial())

summ(mstep3_persist, exp = TRUE, model.fit = FALSE, model.info = FALSE, digits = 3)

```

Interpretation of control predictor `enrollment_units`: 

**Response:** _______________________

Significance of predictor estimate: 

**Response:** _______________________

How did the addition of the control variable to the model affect the focal predictor est/sig? 

**Response:** _______________________


------------------------------------------------------------------------

### GROUP 4: Add the control variable `financial_aid` to the model

**Instructions:**

1. Estimate the following logistic regression model:

*dropped_out = ⍺ + β1(offcampus_work_hours) + β2(cumulative_gpa) + β3(enrollment_units) + β4(financial_aid)*

2. Interpret the Odds Ratio of the predictor `financial_aid` as percent change

**OR-to-Percent-Change Equation:** Percent Change = (1 - OR)*100

3. Is the predictor estimate for `financial_aid` significant? 

4. How did the addition of the control variable to the model affect the focal predictor est/sig? 

------------------------------------------------------------------------

```{r}
mstep4_persist <- glm(dropped_out ~ 
                   offcampus_work_hours +
                   cumulative_gpa +
                   enrollment_units +
                   financial_aid,
                 data = cc_data, family = binomial())

summ(mstep4_persist, exp = TRUE, model.fit = FALSE, model.info = FALSE, digits = 3)

```

Interpretation of control predictor `financial_aid`: 

**Response:** _______________________

Significance of predictor estimate: 

**Response:** _______________________

How did the addition of the control variable to the model affect the focal predictor est/sig? 

**Response:** _______________________


------------------------------------------------------------------------

### Do Adverse Childhood Experiences (ACEs) increase likelihood of commiting a violent crime?

> Data was simulated to approximately replicate findings from study:

**Kazeem et al., 2020**: Study Population (Nigeria): Incarcerated (N=253), Non-incarcerated (N=253)

------------------------------------------------------------------------

Read in the simulated data & specify factors (& factor levels) of categorical variables: 
```{r}

ace_data <- read_csv(here("data", "ACEs_violent_crime_SIM.csv")) |>
  mutate(
    group = factor(group, levels = c("non_incarcerated","incarcerated")),
    sex = factor(sex),
    education_level = factor(education_level,
      levels = c("no_school","<primary","primary","secondary","college","postgrad"))) %>%
  mutate(violent_crime = factor(violent_crime, labels = c("No", "Yes"))) %>% 
  select(violent_crime, ace_family_env, ace_peer_violence, ace_community_violence,
         ace_war_exposure, ses_low, age, education_level, sex, ethnicity)

```

Plot the categorical outcome & view response proportions for each category:
```{r}

bar_plot(
  data = ace_data,
  x_var = violent_crime,
  x_label = "Participant Committed a Violent Crime",
  title = "Categorical Outcome Variable Violent Crime") + 
  labs(subtitle = "Sample Population (Nigeria): Incarcerated (N=253), Non-incarcerated (N=253)")

```

------------------------------------------------------------------------

### GROUP 5: Estimate the regression of the outcome predicted by focal predictor

**Instructions:**

1. Estimate the following logistic regression model:

*violent_crime = ⍺ + β1(ace_family_env)*

2. Convert the intercept from Odds-Ratio (OR) to probability and interpret:

**OR-to-Probability Equation:**  Probability = OR_intercept / (1 + OR_intercept)

3. Interpret the Odds Ratio of the predictor `ace_family_env` as percent change

**OR-to-Percent-Change Equation:** Percent Change = (1 - OR)*100

4. Is the predictor estimate for `ace_family_env` significant? 

------------------------------------------------------------------------

```{r}
mstep1_aces <- glm(
  violent_crime ~ ace_family_env,
  data = ace_data,
  family = binomial())

summ(mstep1_aces, exp = TRUE, model.fit = FALSE, model.info = FALSE, digits = 3)

```

Interpretation of intercept (probability): 

**Response:** _______________________

Interpretation of focal predictor: 

**Response:** _______________________

Significance of predictor estimate: 

**Response:** _______________________


------------------------------------------------------------------------

### GROUP 6: Add the control variable `ace_peer_violence` to the model

**Instructions:**

1. Estimate the following logistic regression model:

*violent_crime = ⍺ + β1(ace_family_env) + β2(ace_peer_violence)*

2. Interpret the Odds Ratio of the predictor `ace_peer_violence` as percent change

**OR-to-Percent-Change Equation:** Percent Change = (1 - OR)*100

3. Is the predictor estimate for `ace_peer_violence` significant? 

4. How did the addition of the control variable to the model affect the focal predictor est/sig? 

------------------------------------------------------------------------

```{r}
mstep2_aces <- glm(
  violent_crime ~ ace_family_env + ace_peer_violence,
  data = ace_data,
  family = binomial())

summ(mstep2_aces, exp = TRUE, model.fit = FALSE, model.info = FALSE, digits = 3)

```

Interpretation of control predictor: 

**Response:** _______________________

Significance of predictor estimate: 

**Response:** _______________________

How did the addition of the control variable to the model affect the focal predictor est/sig? 

**Response:** _______________________

------------------------------------------------------------------------

### GROUP 7: Add the control variable `ace_community_violence` to the model

**Instructions:**

1. Estimate the following logistic regression model:

*violent_crime = ⍺ + β1(ace_family_env) + β2(ace_peer_violence) + β3(ace_community_violence)*

2. Interpret the Odds Ratio of the predictor `ace_community_violence` as percent change

**OR-to-Percent-Change Equation:** Percent Change = (1 - OR)*100

3. Is the predictor estimate for `ace_community_violence` significant? 

------------------------------------------------------------------------

```{r}
mstep3_aces <- glm(
  violent_crime ~ 
    ace_family_env + ace_peer_violence + ace_community_violence,
  data = ace_data,
  family = binomial())

summ(mstep3_aces, exp = TRUE, model.fit = FALSE, model.info = FALSE, digits = 3)

```

Interpretation of control predictor: 

**Response:** _______________________

Significance of predictor estimate: 

**Response:** _______________________

------------------------------------------------------------------------

### GROUP 8: Add the control variable `ace_war_exposure` to the model

**Instructions:**

1. Estimate the following logistic regression model:

*violent_crime = ⍺ + β1(ace_family_env) + β2(ace_peer_violence) + β3(ace_community_violence) + β4(ace_war_exposure)*

2. Interpret the Odds Ratio of the predictor `ace_war_exposure` as percent change

**OR-to-Percent-Change Equation:** Percent Change = (1 - OR)*100

3. Is the predictor estimate for `ace_war_exposure` significant? 

------------------------------------------------------------------------

```{r}
mstep4_aces <- glm(
  violent_crime ~ 
  ace_family_env + ace_peer_violence + ace_community_violence + ace_war_exposure,
  data = ace_data,
  family = binomial())

summ(mstep4_aces, exp = TRUE, model.fit = FALSE, model.info = FALSE, digits = 3)

```

Interpretation of control predictor: 

**Response:** _______________________

Significance of predictor estimate: 

**Response:** _______________________

------------------------------------------------------------------------

### GROUP 9: Add the control variable `ses_low` to the model

**Instructions:**

1. Estimate the following logistic regression model:

*violent_crime = ⍺ + β1(ace_family_env) + β2(ace_peer_violence) + β3(ace_community_violence) + β4(ace_war_exposure) + β5(ses_low)*

2. Interpret the Odds Ratio of the predictor `ses_low` as percent change

**OR-to-Percent-Change Equation:** Percent Change = (1 - OR)*100

3. Is the predictor estimate for `ses_low` significant? 

------------------------------------------------------------------------

```{r}
mstep5_aces <- glm(
  violent_crime ~ 
    ace_family_env + ace_peer_violence + ace_community_violence + ace_war_exposure +
    ses_low,
  data = ace_data,
  family = binomial())

summ(mstep5_aces, exp = TRUE, model.fit = FALSE, model.info = FALSE, digits = 3)

```

Interpretation of control predictor: 

**Response:** _______________________

Significance of predictor estimate: 

**Response:** _______________________

------------------------------------------------------------------------

### GROUP 10: Add the control variable `sex` to the model

**Instructions:**

1. Estimate the following logistic regression model:

*violent_crime = ⍺ + β1(ace_family_env) + β2(ace_peer_violence) + β3(ace_community_violence) + β4(ace_war_exposure) + β5(ses_low) + β6(sex)* 

2. Interpret the Odds Ratio of the predictor `sex` as percent change

**OR-to-Percent-Change Equation:** Percent Change = (1 - OR)*100

3. Is the predictor estimate for `sex` significant? 

------------------------------------------------------------------------

```{r}
mstep6_aces <- glm(
  violent_crime ~ 
    ace_family_env + ace_peer_violence + ace_community_violence + ace_war_exposure +
    ses_low + 
    sex,
  data = ace_data,
  family = binomial())

summ(mstep6_aces, exp = TRUE, model.fit = FALSE, model.info = FALSE, digits = 3)

```

Interpretation of control predictor: 

**Response:** _______________________

Significance of predictor estimate: 

**Response:** _______________________


------------------------------------------------------------------------

------------------------------------------------------------------------